---
title: "Practical Machine Learning Project"
author: "Marvin Zaluski"
date: "December 1, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(knitr)
library(plyr)
library(caret)
library(scales)

```

## Data Description

``` {r dd01, echo=F, message=F, warning=F}

train.raw <- read.csv("data/pml-training.csv")
test.raw <- read.csv("data/pml-testing.csv")

train.class.count <- count(train.raw, var="classe")
train.class.count$percent <- percent(train.class.count$freq / sum(train.class.count$freq)) 

train.na <- apply(train.raw, 2, FUN=function(x) which(is.na(x)))
train.na.count <- lapply(train.na, FUN=function(x) length(x))
train.na.cols <- train.raw[, which(train.na.count>0)]
train.na.cols.na <- apply(train.na.cols, 1, FUN=function(x) which(is.na(x)))
train.na.cols.na.count <- lapply(train.na.cols.na, FUN=function(x) length(x))

```

The objective of this report is to describe a predictive model used in identify the way people exercise. The data used in developing the model was collected from personal devices such as Jawbone up, Nike Fuelband, and Fitbit. This data set contains `r ncol(test.raw)` variables used in the modeling process. The training data set contains `r format(nrow(train.raw), big.mark=",")` observations and the test has `r nrow(test.raw)`. The following characteristics are derived from the training data set:

- comment on the different types of data
- the distribution of exercise types is seen below. They are close to evenly distributed with the majority in class Exercise `r train.class.count[train.class.count$freq == max(train.class.count$freq), "classe"]` and minority to class Exercise `r train.class.count[train.class.count$freq == min(train.class.count$freq), "classe"]`. 
`r kable(train.class.count, caption="Class Distribution", format.args=list(big.mark=","), col.names=c("Class", "Number", "Percentage"))`
- some of the following variables have a large amount of missing values (`r percent(round(unique(train.na.count[train.na.count>0])[[1]] / nrow(train.raw),2))`) such as `r paste(names(train.na.cols)[1:3],sep=",")`, and `r names(train.na.cols)[4]`. The rows with incomplete data have the same variables missing for each.
- comment on the variance for some of the variables
- comment on the distribution of classes for discrete variables.


## Data Preprocessing

### Filtering

### Transformation

### Combination

## Data Modeling

``` {r dm01, echo=F, message=F, warning=F}


```

## Results and Discussion

## Conclusion
